Categorical Predictor Embeddings in Actuarial Modeling

### 2021-04-04 manuscript/manuscript.Rmd

These model architectures are relatively uncomplicated by today's standards, and are so chosen to better highlight the
embedding components. Later, in Section \ref{attn}, we will investigate more involved architectures utilizing embeddings
that represent the state of the art for modeling tabular data.

@@

In both approaches, the overarching theme is as follows: We have a collection of points in higher dimensional space,
$\mathcal{X} = \{x_1, x_2, \dots, x_n\}$ that we wish to map to some lower dimensional space, typically $\mathbb{R}^2$,
obtaining $\mathcal{Y} = \{y_1, y_2, \dots, y_n\}$. PCA performs a linear change of coordinates on 
the input data so that the variance of the data is maximized along each orthogonal coordinate, also known as
*principal component*, successively. Formally, if we denote $\mathbf{X}$ and $\mathbf{Y}$ to be the matrices whose 
columns are elements of 
$\mathcal{X}$ and $\mathcal{Y}$, respectively, this can be represented as looking for an orthonormal matrix $\mathbf{P}$
in $\mathbf{Y} = \mathbf{PX}$ such that the covariance matrix of $\mathbf{Y}$, $\frac{1}{n}\mathbf{YY^T}$, is diagonal.
Here, the rows of $\mathbf{P}$ are the principal components that define the resulting coordinate system. It turns out
that that these principal components are the eigenvectors of the covariance matrix of $\mathbf{X}$, i.e., 
$\mathbf{C_X} \equiv \frac{1}{n}\mathbf{XX^T}$. In practice, numerical procedures, including R's \texttt{prcomp()}
which we use, utilize singular value decomposition (SVD) to compute the eigenvectors.

@@

```{r flood-zone-pca, echo = FALSE, fig.cap="Flood zone PCA", out.width = "50%", fig.align = "center"}
knitr::include_graphics("images/flood_zone_pca.png")
```

@@

Let $p_{j|i}$ denote the probability that point $x_i$ picks point $x_j$ as its *neighbor*, if it picks neighbors in
proportion to their density under an $n$-dimension Gaussian centered at $x_i$. Formally, we can write $p_{j|i}$ as

\begin{equation}
p_{j|i} = \frac{\exp(\norm{x_i - x_j}^2 / 2\sigma_i^2)}{\sum_{k\neq i} \exp(-\norm{x_i - x_k}^2 / 2\sigma_i^2)}. (\#eq:pji)
\end{equation}

The similarity metric for the originals space symmetrizes this conditional probability and is defined as

@@

While t-SNE is a powerful visualization technique, there are a few points the modeler should be mindful of in practice.
We refer the reader to @wattenberg2016how for a more thorough discussion, but stress here that the resulting plot can
be very sensitive to the perplexity hyperparameter. In addition, for our use case, which is to explore the
learned embeddings, we have fewer data points than most other applications in the literature. Accordingly, our
reaosnable range of hyperparameter values need to be adjusted down.
In Figures \@ref(fig:flood-zone-tsne), we display the t-SNE of the learned embeddings for `flood_zone` in
Model 3 with various perplexity settings: 0.1, 0.5, 1.0, and 10; the learning rate and the number of steps are fixed at
100 and 10,000, respectively, for all plots. In the first three plots, we can see the semblance of clusters forming. In
particular, flood zone D appears in its own cluster, which implies it is far away from other levels in the embedding
space. When the perplexity is set to 10 (and beyond), however, we lose the geometric properties.

```{r flood-zone-tsne, echo = FALSE, fig.cap="Flood zone t-SNE", out.width = "60%", fig.align = "center"}
knitr::include_graphics("images/flood_zone_tsne.png")
```

@@

While linear regression (i.e., the specific case of GLM with a gaussian distribution and identity link function) is not
commonly used for severity modeling, we throw it into the mix as a benchmark, a practice we recommend in order to
check reasonablness of other models' results. Since it is possible for a linear
regression to output negative values, we cap the predictions below by 0.01 during scoring. Somewhat surprisingly, in 
the problem we
are considering, it performs similary to the more complex neural network models in the RMSE metric (which it optimizes
for, as the unit deviance of the normal distribution corresponds with MSE.) Of course, this is not to discredit neural
networks or GLMs, as the experiments are performed without extensive hyperparameter tuning or feature engineering.

>>>>

These model architectures are relatively uncomplicated by today's standards, and are so chosen to better highlight the
embedding components. Later, in Section \ref{attention_models}, we will investigate more involved architectures utilizing embeddings
that represent the state of the art for modeling tabular data.

@@

In both approaches, the overarching theme is as follows: we have a collection of points in higher dimensional space,
$\mathcal{X} = \{x_1, x_2, \dots, x_n\}$ that we wish to map to some lower dimensional space, typically $\mathbb{R}^2$,
obtaining $\mathcal{Y} = \{y_1, y_2, \dots, y_n\}$. PCA performs a linear change of coordinates on 
the input data so that the variance of the data is maximized along each orthogonal coordinate, also known as
*principal component*, successively. Formally, if we denote $\mathbf{X}$ and $\mathbf{Y}$ to be the matrices whose 
columns are elements of 
$\mathcal{X}$ and $\mathcal{Y}$, respectively, this can be represented as looking for an orthonormal matrix $\mathbf{P}$
in $\mathbf{Y} = \mathbf{PX}$ such that the covariance matrix of $\mathbf{Y}$, $\frac{1}{n}\mathbf{YY^T}$, is diagonal.
Here, the rows of $\mathbf{P}$ are the principal components that define the resulting coordinate system. It turns out
that that these principal components are the eigenvectors of the covariance matrix of $\mathbf{X}$, i.e., 
$\mathbf{C_X} \equiv \frac{1}{n}\mathbf{XX^T}$. In practice, numerical procedures, including R's \texttt{prcomp()}
which we use, utilize singular value decomposition (SVD) to compute the eigenvectors.

@@

```{r flood-zone-pca, echo = FALSE, fig.cap="First two principal components of learned embeddings of flood zone in Model 4.", out.width = "50%", fig.align = "center"}
knitr::include_graphics("images/flood_zone_pca.png")
```

@@

Let $p_{j|i}$ denote the probability that point $x_i$ picks point $x_j$ as its *neighbor*, if it picks neighbors in
proportion to its density under an $n$-dimension Gaussian centered at $x_i$. Formally, we can write $p_{j|i}$ as

\begin{equation}
p_{j|i} = \frac{\exp(\norm{x_i - x_j}^2 / 2\sigma_i^2)}{\sum_{k\neq i} \exp(-\norm{x_i - x_k}^2 / 2\sigma_i^2)}, (\#eq:pji)
\end{equation}
where $\sigma_i^2$ denotes the variance of the Gaussian associated with $x_i$. The similarity metric for the original
 space symmetrizes this conditional probability and is defined as

@@

While t-SNE is a powerful visualization technique, there are a few points the modeler should be mindful of in practice.
We refer the reader to @wattenberg2016how for a more thorough discussion, but stress here that the resulting plot can
be very sensitive to the perplexity hyperparameter. In addition, for our use case, which is to explore the
learned embeddings, we have fewer data points than most other applications in the literature. Accordingly, our
reaosnable range of hyperparameter values need to be adjusted down.
In Figures \@ref(fig:flood-zone-tsne), we display the t-SNE of the learned embeddings for `flood_zone` in
Model 3 with various perplexity settings: 2, 3, 5, and 10; the learning rate and the number of steps are fixed at
100 and 10,000, respectively, for all plots. In the first three plots, we can see the semblance of clusters forming. In
particular, flood zone D appears in its own cluster, which implies it is far away from other levels in the embedding
space. When the perplexity is set to 10 (and beyond), however, we start to lose the geometric properties.

```{r flood-zone-tsne, echo = FALSE, fig.cap="t-SNE plots of learned embeddings of flood zone in Model 4, with various perplexity values. From left to right then top to bottom: 2, 3, 5, and 10.", out.width = "60%", fig.align = "center"}
knitr::include_graphics("images/flood_zone_tsne.png")
```

@@

While linear regression (i.e., the specific case of GLM with a gaussian distribution and identity link function) is not
commonly used for severity modeling, we throw it into the mix as a benchmark, a practice we recommend in order to
check reasonablness of other models' results. Since it is possible for a linear
regression to output negative values, we cap the predictions below by 0.01 during scoring. Somewhat surprisingly, in 
the problem we
are considering, it performs similary to the more complex neural network models in the RMSE metric (which it optimizes
for, as the unit deviance of the normal distribution correspond to MSE.) Of course, this is not to discredit neural
networks or GLMs, as the experiments are performed without extensive hyperparameter tuning or feature engineering.
